---
title: "Milestone 3 Group 10"
output:
  html_document: default
  pdf_document: default
date: "2024-11-10"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##FILE IMPORTATION

#We exported cleaned data sets for medicare, medicaid, mortality and
population data as csv files into our Project 251 folder from Milestone
2 Rmd file. We shall now import the medicare, mortality and population
cleaned data sets into the Milestone 3 Rmd file for preparation for
further analysis.

#We have chosen to use the medicare data (65+) to be prepared for
merging with mortality and population data sets.

```{r}
library(readr)
setwd("/home/rstudio/Project251/milestone_3/data_sets_for_milestone_3")

medicare_data <- read_csv('medicare_data_cleaned.csv')
cdc_wonder_mortality_data <- read_csv("cdc_wonder_mortality_data_cleaned.csv")
cdc_wonder_population_data <- read_csv("cdc_wonder_population_data_cleaned.csv")


head(medicare_data)
head(cdc_wonder_mortality_data)
head(cdc_wonder_population_data)

```

##RECODING MEDICARE_DATA

#Renamed the first column to residence_state as is in the mortality data
sets, select the first three columns with totals on enrolemnt,
utilisation and payments, and remove the "UNITED STATES" row with
totals. We have capitalised each word in the state name and converted
them to lower case as is the case in the mortality dataset.

#Removed dollar signs and commas from the program payment using the gsub
function, then converted the column variables to numeric.

#Finally, CREATED A NEW VARIABLE, "medicare_spending_per_person", is
created by dividing the total program payments by the number of
enrollees to represent average spending per enrollee for each state.

```{r}
library(stringr)
library(dplyr)

medicare_data_modified <- medicare_data %>% 
  select(residence_state = location, 
         total_traditional_medicare_part_a_and_or_part_b_enrollees, 
         total_medicare_part_a_and_or_part_b_persons_with_utilization, 
         total_medicare_part_a_and_or_part_b_program_payments) %>% 
  filter(residence_state != 'UNITED STATES') %>%
  mutate(
    residence_state = str_to_title(tolower(residence_state)), 
    residence_state = ifelse(residence_state == "District Of Columbia", "District of Columbia", residence_state)
  )

medicare_data_modified <- medicare_data_modified %>%
  mutate(total_medicare_part_a_and_or_part_b_program_payments = 
           as.numeric(gsub("[\\$,]", "", total_medicare_part_a_and_or_part_b_program_payments)))

medicare_data_modified <- medicare_data_modified %>%
  mutate(medicare_spending_per_person = total_medicare_part_a_and_or_part_b_program_payments / total_traditional_medicare_part_a_and_or_part_b_enrollees)

head(medicare_data_modified)

str(medicare_data_modified)
```

##RECODING FOR CDC_WONDER_MORTAILITY_DATA

#Filtering and Selecting Columns:

#We filtered the cdc_wonder_mortality_data to keep only records for age
groups 65-74 years, 75-84 years, and 85+ years. We then selected only
the columns residence_state, ten_year_age_groups, and deaths.

#Aggregating Deaths: We grouped the filtered data by residence_state and
ten_year_age_groups. Using summarize(), we calculated the total number
of deaths for each group combination, storing the result in a new
column, total_deaths. The .groups = 'drop' argument ensures that the
grouping is dropped after summarizing.

```{r}
mortality_modified <- cdc_wonder_mortality_data %>% 
  filter(ten_year_age_groups %in% c('65-74 years', '75-84 years', '85+ years')) %>% 
  select(residence_state, ten_year_age_groups, deaths)


mortality_modified<- mortality_modified %>% 
  group_by(residence_state, ten_year_age_groups) %>% 
  summarize(total_deaths = sum(deaths), .groups = 'drop')

head(mortality_modified)
str(mortality_modified)


```

#Filtering by Relevant Age Group (65+). Since Medicare data focuses on
the 65+ population, we ensured that the mortality data contains an age
column that allows filtering for this age group. we created a new
variable "total_deaths_65_plus" to represent the total deaths 65+ per
state. We performed this using the summarize () for the ten year age
groups after grouping by residence state.

```{r}
mortality_65_plus <- mortality_modified %>% 
  group_by(residence_state) %>% 
  summarize(total_deaths_65_plus = sum(total_deaths), .groups = 'drop')

head(mortality_65_plus)
```

##RECODING FOR CDC POPULATION DATA

#Standardizing the state column name to ensure that the column
representing states is named residence_state to match the other
datasets. using the rename function.

#Removed prefixes from residence state names e.g., 01-Alabama using gsub
function.

#Filtering by Relevant Age Group (65+). Since Medicare data focuses on
the 65+ population, we will ensure that the population data contains an
age column that allows filtering for this age group.we shall create a
new variable "total_population_65_plus" to represent the total
population 65+ per state. We performed this using the summarize function
for the ten year age groups after grouping by residence state.

```{r}

population_data_modified<- cdc_wonder_population_data %>%
   rename(residence_state = state)    %>%
mutate(residence_state = gsub("^[0-9]+-", "", residence_state)) %>%
  filter(ten_year_age_groups %in% c('65-74 years', '75-84 years', '85+ years'))

population_65_plus <- population_data_modified %>%
group_by(residence_state) %>% 
  summarize(total_population_65_plus = sum(population), .groups = 'drop')


head(population_65_plus)
str(population_65_plus)
```

##CREATING A NEW VARIABLE; MORTALITY RATE PER 1000 POPULATION

#To create the mortality_rate_per_1000 variable, we divided the
total_deaths_65_plus by the population count for each state and then
multiply by 1,000 to get the rate per 1,000 people. We first merged the
mortality_65_plus dataset with the population_65_plus dataset using
left_join () and then calculated the mortality rate.

```{r}
mortality_population_merged <- mortality_65_plus %>%
  left_join(population_65_plus, by = "residence_state")

mortality_rate <- mortality_population_merged %>%
  mutate(mortality_rate_per_1000 = (total_deaths_65_plus / total_population_65_plus) * 1000)

head(mortality_rate)
str(mortality_rate)
```

#DESCRIPTIVE STATISTICS #We calculated descriptive statistics, including
mean, median, and range, for key variables in the merged dataset
(total_deaths_65_plus, total_population_65_plus, and
mortality_rate_per_1000). Using summarize() with mean(), median(), and
range(), we obtained an overview of the central tendencies and spread of
these variables. This provides insights into the distribution of deaths,
population, and mortality rates across states for individuals aged 65
and older.

```{r}
library(dplyr)

descriptive_stats <- mortality_rate %>%
  summarize(
    mean_total_deaths = mean(total_deaths_65_plus, na.rm = TRUE),
    median_total_deaths = median(total_deaths_65_plus, na.rm = TRUE),
    range_total_deaths = range(total_deaths_65_plus, na.rm = TRUE),
    
    mean_total_population = mean(total_population_65_plus, na.rm = TRUE),
    median_total_population = median(total_population_65_plus, na.rm = TRUE),
    range_total_population = range(total_population_65_plus, na.rm = TRUE),
    
    mean_mortality_rate_per_1000 = mean(mortality_rate_per_1000, na.rm = TRUE),
    median_mortality_rate_per_1000 = median(mortality_rate_per_1000, na.rm = TRUE),
    range_mortality_rate_per_1000 = range(mortality_rate_per_1000, na.rm = TRUE)
  )

descriptive_stats

```

#DATA DICTIONARY

```{r echo=FALSE, message=FALSE}
library(readxl)
library(knitr)

data_dictionary <- read_excel("data_dictionary_project_milestone_3.xlsx")

kable(data_dictionary, caption = "Data Dictionary for Milestone 3 Group 10")

```

###SUMMARY OF OUR WORK

##Subset Rows or Columns as Needed Rubric Requirement: Subset the
dataset to include only relevant rows and columns.

#Action: We have subset columns to include only those relevant to
analysis for Medicare, mortality, and population datasets.

##Create New Variables Needed for Analysis Rubric Requirement: Create at
least two new variables that are useful for the analysis.

#Actions: A mortality_rate_per_1000 variable by dividing total deaths by
the population for each state and then multiplying by 1,000. A
medicare_spending_per_person variable by dividing total Medicare
spending by the number of enrollees for each state. These variables will
help in analyzing spending vs. mortality rates more effectively.

##Clean Variables Needed for Analysis Rubric Requirement: Clean at least
two variables to ensure consistency and usability.

#Actions: We have standardized state names, converting spending amounts
to numeric, and filtering to the relevant age group (65+). All relevant
variables (spending, population, mortality counts) are numeric and
formatted consistently. The column names are now clear and consistent
across datasets.

##Restructure Datasets as Needed Rubric Requirement: Restructure
datasets, if necessary, to prepare for analysis or merging.

Action: We've prepared each dataset for merging by standardizing column
names (residence_state).

##Data Dictionary Based on Clean Dataset Rubric Requirement: Create a
data dictionary with at least four data elements from the cleaned
dataset, describing each variable's meaning, type, and units (if
applicable).

#Actions: We created a table listing variables such as residence_state,
total_deaths, mortality_rate, and medicare_spending, with descriptions
of each variable, data types, and units. This was done in an excel file
and imported into the Rmd file and later knitted to an html document.

##One or More Tables with Descriptive Statistics for 4 Data Elements
Rubric Requirement: Present descriptive statistics (mean, median, range,
etc.)

#Actions: Used summarize() to create a table with descriptive statistics
for: o medicare_spending_per_person o mortality_rate_per_1000 o
total_population_65_plus o total_deaths_65_plus

##HTML Document That is Professionally Prepared for Presentation Rubric
Requirement: Create an HTML document that presents the analysis in a
professional manner.

#Action: Used R Markdown to compile all code, tables, and visualizations
into a professional HTML document. Ensured it has clear sections,
headings, and explanations of the analysis steps.
